Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2023-03-13 intro.tex
gravity negligible at colliders
P_{ba} -> P_{ab}

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations of quantum fields. The symmetry group
    of the SM is
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $c$ describes the interactions of the
    strong force within the theory of quantum chromodynamics, which we
    will elaborate more on in Section~\ref{sec:qcd}. The second
    and third gauge groups represent the electroweak sector of the SM, where
    $L$ denotes left-chiral fields that carry weak isospin, whilst $Y$ denotes
    the weak hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces that we observe.
    The explanation of gravity in a QFT is an open problem and cannot currently
    be included in the SM as any Lagrangian including gravity cannot be renormalised.
    Fortunately, the effect of gravity is considered
    to be negligible on the scales considered in the SM, and so it is ignored.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The massless gauge bosons mediating the strong
    and EM forces are the gluon $g$ and photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons, which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, elucidating the Higgs boson $H$.

@@

    For a review of PDFs and how they are determined
    see Refs. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs},
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(z,\mu_{F}) \\
            f_{g/h}(z,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\frac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$, that carries
    a momentum fraction $x$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops \cite{Moch:2004pa,Vogt:2004mw}.
    At leading order they are \cite{Altarelli:1977zs}
    \begin{equation}\label{eqn:AP_kernels}
        \begin{split}
            P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \, , \\
            P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \, ,\\
            P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \, ,\\
            P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z) \, ,
        \end{split}
    \end{equation}
    where the divergences at $z=1$ have been regulated
    with the `$+$'-prescription, which is defined as
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, .
    \end{equation}
    With this prescription, the divergence at $g(z=1)$ is cancelled,
    given that the function $f(z)$ is sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a momentum fraction $x$ of the parent parton.

>>>>

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations of quantum fields. The symmetry group
    of the SM is
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $c$ describes the interactions of the
    strong force within the theory of quantum chromodynamics, which we
    will elaborate more on in Section~\ref{sec:qcd}. The second
    and third gauge groups represent the electroweak sector of the SM, where
    $L$ denotes left-chiral fields that carry weak isospin, whilst $Y$ denotes
    the weak hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces that we observe.
    The explanation of gravity in a QFT is an open problem and cannot currently
    be included in the SM as any Lagrangian including gravity cannot be renormalised.
    Fortunately, the effect of gravity is considered
    to be negligible on the scales considered at high energy colliders, and so it is ignored.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The massless gauge bosons mediating the strong
    and EM forces are the gluon $g$ and photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons, which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, elucidating the Higgs boson $H$.

@@

    For a review of PDFs and how they are determined
    see Refs. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs},
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(z,\mu_{F}) \\
            f_{g/h}(z,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ab}(\frac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$, that carries
    a momentum fraction $x$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops \cite{Moch:2004pa,Vogt:2004mw}.
    At leading order they are \cite{Altarelli:1977zs}
    \begin{equation}\label{eqn:AP_kernels}
        \begin{split}
            P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \, , \\
            P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \, ,\\
            P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \, ,\\
            P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z) \, ,
        \end{split}
    \end{equation}
    where the divergences at $z=1$ have been regulated
    with the `$+$'-prescription, which is defined as
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, .
    \end{equation}
    With this prescription, the divergence at $g(z=1)$ is cancelled,
    given that the function $f(z)$ is sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a momentum fraction $x$ of the parent parton.
